Abstractions for Multi-Sorted Substitutions

### 2024-03-11 paper/paper.tex
Update.

  \begin{figure}
    \label{fig:map-ops}
    \centering
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KMap
    \LibCode\KAp
    \LibCode\KLiftAlt
    \LibCode\KSingleAlt
    \end{minipage}
    \begin{minipage}[t]{0.4\linewidth}
    \LibCode\KWeakenAlt
    \LibCode\KEq
    \LibCode\KFunExt
    \LibCode\KId
    \LibCode\KIdLift
    \end{minipage}
    \caption{Map Operations}
  \end{figure}
  Figure~\ref{fig:map-ops} shows the usual operations for renamings and substitutions
  generalized to maps.\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  The type \ACode{S₁ \ADef{→ₖ} S₂} unifies
    renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}.
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  The operation \ACode{ϕ \& x} applies a map to a variable.
  The operation \ACode{ϕ ↑ s} lifts a map under a binder of sort \ACode{s}.
  The operation \ACode{⦅ x/t ⦆} constructs a singleton map that replaces
  \ACode{\ACon{zero}} with \ACode{x/t} and decreases all other variables by one.
  The \ACode{\AField{weaken s}} map increases all variables by one.
  The \ACode{ϕ₁ $\sim$ ϕ₂} type expresses extensional equality of maps.
    For simplicity, we postulate functional extensionality \ACode{\~-ext}.\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
  There is an identity map \ACode{id}.
  The lemma \ACode{id↑-id} states that a lifted identity map is again an identity map.

@@

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the axioms of a \ACode{\ACon{CTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeaken
    % \LibCode\KCommLiftWeakenProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    statement for multiple map applications:
    \LibCode\KCommLiftWeakenTraverse
    % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingle
    % \LibCode\KWeakenCancelsSingleProof
    By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    this lemma, we get the same statement for multiple map applications:
    \LibCode\KWeakenCancelsSingleTraverse
    % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingle
    % \LibCode\KDistLiftSingleProof
    By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    same statement for multiple map applications:
    \LibCode\KDistLiftSingleTraverse
    % \LibCode\KDistLiftSingleTraverseProof
  \item 
    We define \ACode{\ACon{CKit}} instances for each of the four
    combinations between renamings and substitutions:
    \LibCode\KCKitInstances
    Interestingly, the \ACode{\ADef{Cₛ}} instance has another
    \ACode{\ADef{CKit}} parameter.
    This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
    which yields the \ACode{\ADef{CKit}} between substitutions
    and renamings, and then use this resulting \ACode{\ADef{CKit}}
    as an argument for the case with two substitutions.
    As shown below, Agda's instance resolution is up for the task:

@@

  \noindent
  \begin{minipage}{0.55\linewidth}
    \LibCode\KContextLookupI
  \end{minipage}
  \begin{minipage}{0.45\linewidth}
    \LibCode\KContextLookupII
  \end{minipage}
  % \LibCode\KContextLookup

>>>>

  \begin{figure}
    \centering
    \begin{minipage}[t]{0.5\linewidth}
    \LibCode\KMap
    \LibCode\KAp
    \LibCode\KLiftAlt
    \LibCode\KSingleAlt
    \end{minipage}
    \begin{minipage}[t]{0.4\linewidth}
    \LibCode\KWeakenAlt
    \LibCode\KEq
    \LibCode\KFunExt
    \LibCode\KId
    \LibCode\KIdLift
    \end{minipage}
    \caption{Map Operations}
    \label{fig:map-ops}
  \end{figure}
  Figure~\ref{fig:map-ops} shows the usual operations for renamings and substitutions
  generalized to maps.\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  The type \ACode{S₁ \ADef{→ₖ} S₂} unifies
    renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}.
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map} and
    use the meta-variable \ACode{ϕ} for it.
  The operation \ACode{ϕ \& x} applies a map to a variable.
  The operation \ACode{ϕ ↑ s} lifts a map under a binder of sort \ACode{s}.
  The operation \ACode{⦅ x/t ⦆} constructs a singleton map that replaces
  \ACode{\ACon{zero}} with \ACode{x/t} and decreases all other variables by one.
  The \ACode{\AField{weaken s}} map increases all variables by one.
  The \ACode{ϕ₁ $\sim$ ϕ₂} type expresses extensional equality of maps.
    For simplicity, we postulate functional extensionality \ACode{$\sim$-ext}.\footnote{
      The actual implementation does \emph{not} use any postulates.
      % The workaround is explained in Appendix~\ref{sec:fun-ext}.
    }
  There is an identity map \ACode{id}.
  The lemma \ACode{id↑-id} states that a lifted identity map is again an identity map.

@@

  Similar as for the \ACode{\ACon{Traversal}}, we derive further
  definitions from the axioms of a \ACode{\ACon{CTraversal}}:
  \begin{itemize}
  \item 
    A map \ACode{ϕ} followed by a weakening is equivalent to a
    weakening followed by \ACode{ϕ} that has been lifted over the weakened
    variable:
    \LibCode\KCommLiftWeakenTraverse

    % A map \ACode{ϕ} followed by a weakening is equivalent to a
    % weakening followed by \ACode{ϕ} that has been lifted over the weakened
    % variable:
    % \LibCode\KCommLiftWeaken
    % % \LibCode\KCommLiftWeakenProof
    % By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the same
    % statement for multiple map applications:
    % \LibCode\KCommLiftWeakenTraverse
    % % \LibCode\KCommLiftWeakenTraverseProof
  \item 
    A weakening followed by a singleton substitution cancel each other out:
    \LibCode\KWeakenCancelsSingleTraverse

    % A weakening followed by a singleton substitution cancel each other out:
    % \LibCode\KWeakenCancelsSingle
    % % \LibCode\KWeakenCancelsSingleProof
    % By applying \ACode{\AField{⋯-fusion}} and \ACode{\AField{⋯-id}} to
    % this lemma, we get the same statement for multiple map applications:
    % \LibCode\KWeakenCancelsSingleTraverse
    % % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingleTraverse
    % A singleton map can be swapped with any map \ACode{ϕ}:
    % \LibCode\KDistLiftSingle
    % % \LibCode\KDistLiftSingleProof
    % By applying \ACode{\AField{⋯-fusion}} to this lemma, we get the
    % same statement for multiple map applications:
    % \LibCode\KDistLiftSingleTraverse
    % % \LibCode\KDistLiftSingleTraverseProof
  \item 
    We define \ACode{\ACon{CKit}} instances for each of the four
    combinations between renamings and substitutions:
    \LibCode\KCKitInstances
    Interestingly, the \ACode{\ADef{Cₛ}} instance has another
    \ACode{\ADef{CKit}} parameter.
    This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
    which yields the \ACode{\ADef{CKit}} between substitutions
    and renamings, and then use this resulting \ACode{\ADef{CKit}}
    as an argument for the case with two substitutions.
    As shown below, Agda's instance resolution is up for the task:

@@

  \noindent
  \begin{minipage}[t]{1.0\linewidth}
    \begin{minipage}[t]{0.55\linewidth}
      \LibCode\KContextLookupI
    \end{minipage}
    \begin{minipage}[t]{0.45\linewidth}
      \LibCode\KContextLookupII
    \end{minipage}
  \end{minipage}
  % \LibCode\KContextLookup
